Quality Assessment of OpenStreetMap Footpath Data

### 2020-08-24 report/report.tex
Fix opponent's suggestions

\setlength\parindent{0pt}
\setlength{\parskip}{1em}
\setcounter{tocdepth}{2}

@@

Virtual cities has since long been a popular research topic in the field of computer visualization.
The idea of recreating real world cities in 3D applications has a growing number of applications, including visualization and modeling, urban planning and architecture, as well as media and entertainment in the form of video games and 3D animation.
Traditionally, modeling virtual cities is a labor-intensive and time-consuming task due to the need for manual modeling, texturing and design and placement of individual models in the large 3D scene.
This type of tasks is commonly done by professional 3D graphical artists and designers.
However, in recent years there has been major research work in the field of procedural content generation, meaning automatic or semi automatic generation of 3D models without the direct involvement of a 3D artist.
\textbf{Procedural models} typically involve a machine which algorithmically composes together or alters human-generated 3D assets.
Sometimes a degree of computer-generated randomness is induced into the generative algoritm in order to create a larger and less predictable generative space of 3D models that can be created from the same algorithm.

@@

The field of procedural city reproduction comes with its own host of challenges however.
In practice, urban environments are typically dense and large, spanning a few to hundreds of square kilometers, and are influenced by variables that are hard to quantify, including land policies, government plans, population changes and transportation infrastructure, making it non-trivial to produce 3D city representations with both high accuracy and high detail, while simultaneously having high utility in the intended field of study.
\textbf{(Vanegas et al, 2010)}.

@@

This project will strive to determine the feasibility of procedurally reconstructing a city road network from OSM data by using such models as have been mentioned in this section, focusing on the feasibility of including pedestrian footpaths.
\textbf{Footpaths} include any form of path that is designated for pedestrians, including sidewalks, walkways and trails.
OSM commonly divides city road networks into any of the following 4 categories: arterial (primary) roads, secondary roads, residential roads and footpaths.
For the purpose of this report, a \textbf{road} will be any road whose primary use is human travel, either via a vehicle or by foot.
Note that by this definition even sidewalks are classified as roads, even though they commonly run in connection to larger roads (which are usually reserved for motor vehicles), and thus may not always be thought of as roads in plain language.

@@

When it comes to the quality of available data, it turns out that in practice, geodata databases without errors are rare.
Geodata validation and correction is only an emerging field of research, and many such datasets were usually obtained years ago, and creating error-free datasets was neither possible due to the lack of validation techniques, nor a high priority.
Luckily there are coming more and more sophisticated methods for automatically correcting the most common errors
\textbf{(Biljecki et al, 2016 (2))}.

Early on, at the stage of importing geodata into a procedural city reproduction model, the most critical source of error is the \textbf{positional accuracy} that the data provider can ensure.
Every geodata service has an upper limit on the positional accuracy of their data, which depends primarily on the method of data acquisition that was used.
Different instruments, such as GPS measurements or satellite image stereoscoping, have theoretical upper limits on their precision, which propagate through the data collection and preservation process as a source of positional error
\textbf{(Fan et al, 2014)}.
Specifically in the case of rendering road networks and buildings, a common issue that stems from low positional accuracy is that the 3D models being generated from road and building data might overlap with each other, which can lead to strange-looking mesh intersections and inaccuracies that modelers often want to avoid.
\textbf{Feature overlap} is a type of geometry error that arises due to inaccuracies in the source data, often due to limited precision in the measuring methods used to obtain the data.

@@

To limit visual quirks such as mesh self intersection, it is often in the designer's best interest to have a stable procedure for importing road data that results in minimal feature overlap, while still retaining accuracy to the real world.
A possible way of mitigating this problem is looking towards gathering geodata from volunteer-based geoservices such as OSM.
The OSM project is a freely available wiki-like geodatabase based on Volunteered Geographical Information (VGI), meaning anyone can contribute by adding or editing features.
The OSM world dataset is being consistently updated by volunteers from all over the world, and the main idea is that by using OSM geodata for city reproduction, the temporal aspect of data accuracy would in theory be more assured by the crowd contribution nature.
There are already a few such services that use OSM data to generate 3D geodata visualizations for media and video game applications.
A few prominent examples are Mantle
\footnote{Mantle homepage. Accessed 17-05-2020. https://www.mantle.tech/}
and Mapbox
\footnote{Mapbox homepage. Accessed 17-05-2020. https://www.mapbox.com/}.
Such services are however usually limited in the accuracy and type of data that they can provide.
While they can provide both building footprints and road networks, they usually provide road data with no respect to overlap between features.

@@

This project has dealt with the feasibility of including footpath data when generating 3D representations of OSM data, and preserving the integrity of the 3D road network, avoiding overlap and collision between road meshes.
It was  that most road overlap can be corrected by first detecting critical locations where feature overlap occurs, and subsequently adjust the footpath road widths and individual vertex positions within the margin of the OSM positional accuracy, in order to obtain a slightly altered map with little to no overlap between footpaths and other features.

@@

The positional accuracy will be necessary to carry out the investigation of the first subquestion: the rate of collisions between roads, since it will determine the maximum tolerance in distance between road points before a collision is certain to have occurred.
For the secondary research question: the geodata precision study, the positional accuracy will be estimated by comparing feature points between the OSM building dataset and the Swedish University of Agricultural Sciences (SLU) building dataset with known accuracy. See section 3.4.2 for an overview of the SLU dataset and how it is obtained.
The positional accuracy will be needed to assess where road overlap occurs and whether the collisions can easily be corrected.
For the positional accuracy study, see sections 3.1.1 for a method overview, 3.2 for a walk through of the method, and 4.1.2 for the results.

@@

\begin{enumerate}
    \item The program will not handle terrain features and altitude. It will be assumed that the road mesh can be projected on a flat 2d surface without feature intersection.
    \item It will be assumed that the city map consists of only arterial (primary) roads, secondary roads, residential roads and footpaths.
    \item It will be assumed that the OSM dataset represents all roads as simple polylines, and building footprints as simple connected polygons. Any other features (such as squares, which are roads represented as polygons) will be omitted from the dataset.
    \item The road network is assumed to be a 2D simple graph. This means that any overlapping roads, such as tunnels and bridges that cross over street-level roads, will be eliminated from the dataset. Any self-connected nodes will also be eliminated from the dataset.
    \item All building footprints are assumed to be coherent to their footprints along the z axis. Thus, it will be assumed that no buildings taper or extend beyond their footprints at a higher altitude.
\end{enumerate}

@@

This study will not consider the lineage aspect of the geodata.
Lineage is defined by \textbf{(Van Oort, 2006)} as the historical aspect of the dataset: how it has been obtained, with what instruments and how it has evolved between the moment it was obtained and the moment it was retrieved.
The entire OSM dataset is under constant modification by volunteers, so it is expected that the accuracy will improve over time. Previous studies highlight how the positional accuracy in similar-sized cities have improved over the course of several years (\textbf{Fan et al, 2014}, \textbf{Haklay, 2010}).
The historical aspect of the OSM dataset is available through the OSM History Viewer
\footnote{OSM History Viewer - OpenStreetMap Wiki. Accessed 20-03-2020. https://wiki.openstreetmap.org/wiki/OSM\_History\_Viewer}
, an online debugging tool that lets anyone freely view the change history of individual features in a commit history-like fashion.
There is an option for editors to include a personal note with their changesets to include additional details or motivate why a change was made, but there is no guarantee that the data includes any information on the acquisition method used.

Another aspect of geodata precision is attribute accuracy, which will neither be considered in this study.
Attributes are small flags of metadata that are included with geodata features, letting the user know additional details about a feature that is not reflected in the map, e.g. how many lanes a road has, if the base is asphalt or gravel, or how many stories a building has.
Attribute accuracy is the measure of how complete attribute information is in a dataset.
How much attribute data is missing and how much of it does not conform with reality?
In general, attribute accuracy is hard to quantify.
In this case however, it might be interesting to compare attributes between the OSM map and the SLU reference map that was used in this project, as both use encoded metadata on a per feature basis, and matching features can be easily identified.
See section 3.4.2 for an overview of the reference dataset that was used in this study.
The attribute accuracy of the two datasets could in theory be assessed by doing a simple feature comparison and creating a translation table between the two attribute maps.
To the knowledge of the author, this has never been done with specifically these two datasets, at the time of writing.

@@

In their analysis of procedural city generation models, \textbf{(Müller et al, 2006)} provide an overview of the process of recreating virtual 3D cities from their real life counterparts.
They present cities as complex collections of \textbf{buildings}, \textbf{blocks}, \textbf{parcels} and \textbf{neighborhoods} which are all interconnected via a \textbf{street} network, where individual streets can serve different types of traffic.
These features are commonly represented as \textbf{geodata} in map databases, and processed and organized into visualization-ready data by a \textbf{Geographic Information System (GIS)}.
In commercially available geodata, roads are most often defined as polylines: arrays of global coordinate points that come in the same order as they would normally be visited by a person traveling along the road. Buildings in turn are most often defined by their \textbf{building footprints}, a single polygon with global coordinates as vertices, which defines the extents of the building as seen on the map from above.
In practice, cities are often made up out of many square kilometers worth of geodata, and it would be highly labor intensive to manually recreate all these features in 3D.
Instead, \textbf{procedural models} for city reproduction instead rely on programmatic solutions for converting map data to 3D road and building models, and composing them to a 3D world that the designer may continue to fine-tweak through interfaces for interactive editing.

@@

This project falls within the broader topic of geodata extraction and modeling, primarily of roads and building parcels.
Geodata modeling is at the foundation of procedural modeling of real-world cities, since any software which models cities based on geodata will need a robust method of acquiring and laying out road and building data without compromising the data integrity.
In recent years there has been an advent of interactive tools that aid urban planners in placing or generating features of an urban plan, particularly roads and road networks.
Such tools use procedural methods and AI solutions to move the burden of labor from the human designer to the software that the designer uses
\textbf{(Müller et al, 2006)}.

The last decade has seen multiple startups in the area of procedural software for urban architecture.
Such software uses procedural models to render a city grid based on real-world locations, or generate new grids and features based on user-specified parameters, and allow for varying degrees of manual editing on grid, road, neighborhood, city block or individual feature level.
Some of the most prominent examples are ArcGIS Pro (made by Esri), CityEngine (made by the startup Procedural, bought by ESRI) and Urban Canvas (made by the startup Synthicity, bought by Autodesk, now defunct) \textbf{(Stojanovski et al, 2019)}.

Even as such tools are well available, \textbf{(Stojanovski et al, 2020)} note that they are as of today not commonly applied in urbanist practices, due to failures in reflecting the unique workspace of architects and their design needs.
Urban architecture is a field that not only accounts for spatial data, but also captures many social factors and the individual wills of the many stake holders.
Each architect is an actor in a continuously developing environment, and even the architect themselves is affecting that environment in a complex manner through their actions.

@@

The next major study to assess the quality of OSM data was published by \textbf{(Kunze, 2012)}.
The study applied different methods to assess the completeness of OSM data in two federal states in Germany, mainly by analyzing the area difference between the OSM dataset and an administrative dataset.
Other notable studies about OSM data accuracy were made around the same time in Germany by \textbf{(Zielstra and Zipf, 2010)} and \textbf{(Neis et al, 2012)}, as well as in France by \textbf{(Girres and Touya, 2010)}.

Finally, a major study in 2013-2014, conducted by researchers from various universities in Norway and Germany \textbf{(Fan et al, 2014)}, examined the quality and accuracy in building footprint data of the Munich area, at a time and place where the completeness of the OSM dataset was significantly higher than the study by \textbf{(Haklay, 2010)} (the results of their study showed that the OSM had captured close to 100\% of all building footprints).
This new study included an insight in the geometrical calculations necessary to match features and points between two datasets, and how to reliably calculate the metrics needed for the most important and assessable of van Oort's criteria.
The results of this study revealed that the positional accuracy of the OSM Munich data at the time was about 4 meters

@@

\textbf{(Müller et al, 2006)} present how procedural model generation is accomplished using Shape Grammar: a method developed by \textbf{(Stiny, 1975)} for automatically and iteratively constructing 3D models based in a hierarchical set of rules.
Shape grammar was originally developed from a method known as L-systems: a system for hierarchical model description which was originally used to generate plants for 3D animation \textbf{(Parish and Müller, 2001)}.
Shape grammar works by starting with a crude volumetric model of a building and iteratively remodeling it from a set of shape grammars.
Shape grammars often work from a data bank of 3D models and a set of hierarchical rules for how these can be pieced together.
This creates a possible generative space of many different models that can be created from the same data and rule set, and with the induction of random fluctuations, many different models can be generated from the same dataset and rules.
Using this procedure, a procedural model can structure the façade of a building, add texture and add minor detail models such as windows, doors and ornaments, and so on.
The maximum accuracy that can be obtained in these finer details is largely dependent on the availability of such data, and as most commercially available GIS data only serve crude building footprints and road placement, the model will often have to make guesses about the finer detail, which sacrifices accuracy to the real world.

@@

Space syntax, developed by \textbf{(Hillier, 1997)}, is a set of theories that analyze complex, large-scale spaces such as cities, and tries to explain human behavior from a spatial point of view.
It is commonly used in the analysis of procedural generation models of urban environments.
The central idea is that spatial structures such as building patterns both reflect and create the patterns for human interaction with their surroundings.
A common theme in space syntax is the division of spacial structures into different levels of observation, and then identifying different units of space and the pattern in which they connect to each other.
Space syntax is an analytical method that relies on the availability of city maps, which has since its invention been developed into an extensive research fields on its own
\textbf{(Parish and Müller, 2001)}.
It has also been widely used in the analysis of pedestrian flows and way finding
\textbf{(Peponis, Zimring and Choi, 1990)}.
Furthermore, the idea of using space syntax to divide spatial structures into different levels of observation is frequently used in the research field of urban design and urban planning, to gain an understanding how space is utilized and connected at different levels of usage.
One can for example analyze a city on an interior-by-interior basis, or use city blocks and streets as the least unit of space, depending on the goal of the study.
\textbf{(Stojanovski et al, 2020)}.

@@

To assess the completeness of their OSM data, the study by \textbf{(Fan et al, 2014)} used the fraction of the total building area in the reference dataset and the OSM dataset.
The motivation for this was to eliminate semantic error, which is a type of systematic error due to differences in how data is presented in both datasets.
The most common such difference is the fact that a large building in one dataset may be segmented into several smaller ones in the other.
Because of this, a building in one set might be represented as an aggregation of multiple buildings in the other set (see figure \ref{fig:building-match-types} for a visual example).
This makes it impossible to find a complete one-to-one building footprint mapping between the two datasets, and thus the idea is that using building area as an estimator for completeness will be less sensitive to semantic errors than using the number of buildings or other objects in each dataset.

\textbf{(Fan et al, 2014)} further discussed how the relative overlap in building area can be used as an estimate of the \textbf{building correspondence} between datasets.
Building correspondence is a term that refers to how well it can be judged which building footprints between two datasets that correspond to the same building in real life.
Using the relative area overlap between buildings to find which footprints represent the same building can only be done in cases where there is not much displacement between OSM building footprints data and the reference data set.
The relative building overlap between the OSM and the reference datasets is defined as follows, given the footprints of any building in the OSM set ($foot_{OSM}$) and the reference set ($foot_{REF}$):
\begin{center}
    $S_{RO}(foot_{OSM}, foot_{REF}) = \frac{A_{Overlap}}{min(A_{foot_{OSM}}, A_{foot_{REF}}))}$
\end{center}

@@

Once the building correspondence has been found and a building footprint mapping has been established (where any building in one dataset points to one, more or no buildings in the other dataset) the next challenge is evaluating how similar two matching building footprints are.
When assessing the positional accuracy between the points of two building footprint polygons, first one must find a matching set of vertices between a pair of matching buildings, which is not always trivial since the vertex counts of the two buildings may not be the same.
The two buildings might have been modeled at different levels of detail, or vertex clusters may be found at different parts of the polygon in the two datasets.
To address this problem, the OSM data precision study by \textbf{(Fan et al, 2014)} used the Douglas-Peucker algorithm, first defined by \textbf{(Douglas and Peucker, 1973)}, to reduce both building polygons to a similar vertex count, and then proceed with locating the most likely matching vertices in the polygons.
The full procedure for matching points will be presented in section 3.2.3 of this report.

The study by \textbf{(Fan et al, 2014)} also obtained a measure of the \textbf{shape accuracy} of building footprints.
While the positional accuracy accounts for the error between individual points in a polygon, the shape accuracy is a measure of how similar two polygons at a higher level of observation, regardless of relative scaling and rotation of the two polygons.
The shape accuracy between two matching buildings was quantified by using a similarity function that depends on the difference between the turning functions of matching building footprints in both datasets.
The turning function was first defined by \textbf{(Arkin et al, 1991)}, as a method for measuring the similarity of two polygons, and measures the cumulative angle of a single polygon's counter-clockwise tangent, as a function of the cumulative normalized length when traveling along the edge of the polygon.
The turning function effectively steps over the tangent of a polygon continuously for a full cycle, and logs the total turning angle by adding together the angle of each turn.
It includes consideration of whether the individual turns are clockwise or counter-clockwise, and is invariant to rotation and scaling of the polygon.
Once the turning functions of two similar polygons have been obtained, the integral between them is taken as a measure of shape accuracy.
The full procedure for calculating shape accuracy will be presented in section 3.2.2 of this report.

@@

Another crucial concept in city modeling and GIS mapping is the Level Of Detail (LOD) aspect.
The LOD defines a number of possible different representation of the same GIS data, which vary in the amount of detail in its presentation.
A modeler can choose to present the same data (e.g. building, roads, or decorative models) at different detail levels, depending on the use and application.
Furthermore, the LOD chosen for a particularly application also suggests how the data has been acquired and modeled.
It turns out however, in practice, that the terminology and use of LOD is still ambiguous and the definitions vary greatly between practitioners, standards and institutions, despite its usefulness.
\textbf{(Biljecki et al, 2014)}

@@

The methods used for the geodata precision study were largely based on those used by \textbf{(Fan et al, 2014)} in their study of the quality of the OSM Munich dataslice in 2014.
First a feature map was found, which mapped individual features between the OSM dataset and a reference dataset.
A program then took matching building pairs and extracted matching geometrical points after first simplifying the polygons using the Douglas-Peucker algorithm by \textbf{(Douglas and Peucker, 1973)}.
Once a set of matching points has been found across the whole dataset, the average distance between them was taken as a measure of the positional accuracy.

Additionally, a number of auxiliary calculations were needed to estimate the error boundary of the positional accuracy.
The completeness of the OSM dataset (a measure of how many features are represented in both the OSM dataset and the reference dataset) was calculated as the relative difference in area of all building footprints, between the two datasets.
Additionally, the shape accuracy of the OSM dataset has also been calculated and graphed as an extra verification of the similarity between the datasets.

>>>>

\setlength\parindent{0pt}
\setlength{\parskip}{1em}
\setcounter{tocdepth}{2}
\interfootnotelinepenalty=10000

@@

Virtual cities has since long been a popular research topic in the field of computer visualization.
The idea of recreating real world cities in 3D applications has a growing number of applications, including visualization and modeling, urban planning and architecture, as well as media and entertainment in the form of video games and 3D animation.
Traditionally, modeling virtual cities is a labor-intensive and time-consuming task due to the need for manual modeling, texturing and design and placement of individual models in the large 3D scene.
This type of task is commonly done by professional 3D graphical artists and designers.
However, in recent years there has been major research work in the field of procedural content generation, meaning automatic or semi automatic generation of 3D models without the direct involvement of a 3D artist.
\textbf{Procedural models} typically involve a machine which algorithmically composes together or alters human-generated 3D assets.
Sometimes a degree of computer-generated randomness is induced into the generative algoritm in order to create a larger and less predictable generative space of 3D models that can be created from the same algorithm.

@@

The field of procedural city reproduction comes with its own host of challenges however.
In practice, urban environments are typically dense and large, spanning a few to hundreds of square kilometers, and are influenced by variables that are hard to quantify, including land policies, government plans, population changes and transportation infrastructure, making it non-trivial to produce 3D city representations with both high accuracy and high detail, while simultaneously having high utility in the intended field of study.
\textbf{(Vanegas et al., 2010)}.

@@

This project will strive to determine the feasibility of procedurally reconstructing a city road network from OSM data, focusing on the feasibility of including pedestrian footpaths.
\textbf{Footpaths} include any form of path that is designated for pedestrians, including sidewalks, walkways and trails.
OSM commonly divides city road networks into any of the following 4 categories: arterial (primary) roads, secondary roads, residential roads and footpaths.
For the purpose of this report, a \textbf{road} will be any road whose primary use is human travel, either via a vehicle or by foot.
Note that by this definition even sidewalks are classified as roads, even though they commonly run in connection to larger roads (which are usually reserved for motor vehicles), and thus may not always be thought of as roads in plain language.

@@

When it comes to the quality of available data, it turns out that in practice, geodata databases without errors are rare.
Geodata validation and correction is only an emerging field of research, and many such datasets were obtained years ago, when creating error-free datasets was neither possible due to the lack of validation techniques, nor a high priority.
Luckily there are coming more and more sophisticated methods for automatically correcting the most common errors
\textbf{(Biljecki et al., 2016 (2))}.

Early on, at the stage of importing geodata into a procedural city reproduction model, the most critical source of error is the \textbf{positional accuracy} that the data provider can ensure.
Every geodata service has an upper limit on the positional accuracy of their data, which depends primarily on the method of data acquisition that was used.
Different instruments, such as GPS measurements or satellite image stereoscoping, have theoretical upper limits on their precision, which propagate through the data collection and preservation process as a source of positional error
\textbf{(Fan et al., 2014)}.
Specifically in the case of rendering road networks and buildings, a common issue that stems from low positional accuracy is that the 3D models being generated from road and building data might overlap with each other, which can lead to strange-looking mesh intersections and inaccuracies that modelers often want to avoid.
\textbf{Feature overlap} is a type of geometry error that arises due to inaccuracies in the source data, often due to limited precision in the measuring methods used to obtain the data.

@@

To limit visual quirks such as mesh self intersection, it is often in the designer's best interest to have a stable procedure for importing road data that results in minimal feature overlap, while still retaining accuracy to the real world.
A possible way of mitigating this problem is looking towards gathering geodata from volunteer-based geoservices such as OSM.
The OSM project is a freely available wiki-like geodatabase based on Volunteered Geographical Information (VGI), meaning anyone can contribute by adding or editing features.
The OSM world dataset is being consistently updated by volunteers from all over the world, and the main idea is that by using OSM geodata for city reproduction, the temporal aspect of data accuracy would in theory be more assured by the crowd contribution nature.
There are already a few such services that use OSM data to generate 3D geodata visualizations for media and video game applications.
A few prominent examples are Mantle
\footnote{Mantle homepage. Accessed 17-05-2020. https://www.mantle.tech/}
and Mapbox
\footnote{Mapbox homepage. Accessed 17-05-2020. https://www.mapbox.com/}.
Such services are however usually limited in the accuracy and type of data that they can provide.
While they can provide both building footprints (the polygon that defines the building's extents) and road networks, they usually provide road data with no respect to overlap between features.

@@

This project has dealt with the feasibility of including footpath data when generating 3D representations of OSM data, and preserving the integrity of the 3D road network, avoiding overlap and collision between road meshes.

@@

The positional accuracy will be necessary to carry out the investigation of the second subquestion: the rate of collisions between roads, since it will determine the maximum tolerance in distance between road points before a collision is certain to have occurred.
For the secondary research question: the geodata precision study, the positional accuracy will be estimated by comparing feature points between the OSM building dataset and the Swedish University of Agricultural Sciences (SLU) building dataset with known accuracy. See section 3.4.2 for an overview of the SLU dataset and how it is obtained.
The positional accuracy will be needed to assess where road overlap occurs and whether the collisions can easily be corrected.
For the positional accuracy study, see sections 3.1.1 for a method overview, 3.2 for a walk through of the method, and 4.1.2 for the results.

@@

\begin{enumerate}
    \item The OSM data will be limited to a slice of the Stockholm. See section 3.4 for an exact description of the data that was used used.
    \item The program will not handle terrain features and altitude. It will be assumed that the road mesh can be projected on a flat 2d surface without feature intersection.
    \item It will be assumed that the city map consists of only arterial (primary) roads, secondary roads, residential roads and footpaths.
    \item It will be assumed that the OSM dataset represents all roads as simple polylines, and building footprints as simple connected polygons. Any other features (such as squares, which are roads represented as polygons) will be omitted from the dataset.
    \item The road network is assumed to be a 2D simple graph. This means that any overlapping roads, such as tunnels and bridges that cross over street-level roads, will be eliminated from the dataset. Any self-connected nodes will also be eliminated from the dataset.
    \item All building footprints are assumed to be coherent to their footprints along the z axis. Thus, it will be assumed that no buildings taper or extend beyond their footprints at a higher altitude.
\end{enumerate}

@@

This study will not consider the lineage aspect of the geodata.
Lineage is defined by \textbf{(Van Oort, 2006)} as the historical aspect of the dataset: how it has been obtained, with what instruments and how it has evolved between the moment it was obtained and the moment it was retrieved.
The entire OSM dataset is under constant modification by volunteers, so it is expected that the accuracy will improve over time. Previous studies highlight how the positional accuracy in similar-sized cities have improved over the course of several years (\textbf{Fan et al., 2014}, \textbf{Haklay, 2010}).
The historical aspect of the OSM dataset is available through the OSM History Viewer
\footnote{OSM History Viewer - OpenStreetMap Wiki. Accessed 20-03-2020. https://wiki.openstreetmap.org/wiki/OSM\_History\_Viewer}
, an online debugging tool that lets anyone freely view the change history of individual features in a commit history-like fashion.
There is an option for editors to include a personal note with their changesets to include additional details or motivate why a change was made, but there is no guarantee that the data includes any information on the acquisition method used.

@@

In their analysis of procedural city generation models, \textbf{(Müller et al., 2006)} provide an overview of the process of recreating virtual 3D cities from their real life counterparts.
They present cities as complex collections of \textbf{buildings}, \textbf{blocks}, \textbf{parcels} and \textbf{neighborhoods} which are all interconnected via a \textbf{street} network, where individual streets can serve different types of traffic.
These features are commonly represented as \textbf{geodata} in map databases, and processed and organized into visualization-ready data by a \textbf{Geographic Information System (GIS)}.
In commercially available geodata, roads are most often defined as polylines: arrays of global coordinate points that come in the same order as they would normally be visited by a person traveling along the road. Buildings in turn are most often defined by their \textbf{building footprints}, a single polygon with global coordinates as vertices, which defines the extents of the building as seen on the map from above.
In practice, cities are often made up out of many square kilometers worth of geodata, and it would be highly labor intensive to manually recreate all these features in 3D.
Instead, \textbf{procedural models} for city reproduction instead rely on programmatic solutions for converting map data to 3D road and building models, and composing them to a 3D world that the designer may continue to fine-tweak through interfaces for interactive editing.

@@

This project falls within the broader topic of geodata extraction and modeling, primarily of roads and building parcels.
Geodata modeling is at the foundation of procedural modeling of real-world cities, since any software which models cities based on geodata will need a robust method of acquiring and laying out road and building data without compromising the data integrity.
In recent years there has been an advent of interactive tools that aid urban planners in placing or generating features of an urban plan, particularly roads and road networks.
Such tools use procedural methods and AI solutions to move the burden of labor from the human designer to the software that the designer uses
\textbf{(Müller et al., 2006)}.

The last decade has seen multiple startups in the area of procedural software for urban architecture.
Such software uses procedural models to render a city grid based on real-world locations, or generate new grids and features based on user-specified parameters, and allow for varying degrees of manual editing on grid, road, neighborhood, city block or individual feature level.
Some of the most prominent examples are ArcGIS Pro (made by Esri), CityEngine (made by the startup Procedural, bought by ESRI) and Urban Canvas (made by the startup Synthicity, bought by Autodesk, now defunct) \textbf{(Stojanovski et al., 2019)}.

Even as such tools are well available, \textbf{(Stojanovski et al., 2020)} note that they are as of today not commonly applied in urbanist practices, due to failures in reflecting the unique workspace of architects and their design needs.
Urban architecture is a field that not only accounts for spatial data, but also captures many social factors and the individual wills of the many stake holders.
Each architect is an actor in a continuously developing environment, and even the architect themselves is affecting that environment in a complex manner through their actions.

@@

The next major study to assess the quality of OSM data was published by \textbf{(Kunze, 2012)}.
The study applied different methods to assess the completeness of OSM data in two federal states in Germany, mainly by analyzing the area difference between the OSM dataset and an administrative dataset.
Other notable studies about OSM data accuracy were made around the same time in Germany by \textbf{(Zielstra and Zipf, 2010)} and \textbf{(Neis et al., 2012)}, as well as in France by \textbf{(Girres and Touya, 2010)}.

Finally, a major study in 2013-2014, conducted by researchers from various universities in Norway and Germany \textbf{(Fan et al., 2014)}, examined the quality and accuracy in building footprint data of the Munich area, at a time and place where the completeness of the OSM dataset was significantly higher than the study by \textbf{(Haklay, 2010)} (the results of their study showed that the OSM had captured close to 100\% of all building footprints).
This new study included an insight in the geometrical calculations necessary to match features and points between two datasets, and how to reliably calculate the metrics needed for the most important and assessable of van Oort's criteria.
The results of this study revealed that the positional accuracy of the OSM Munich data at the time was about 4 meters

@@

\textbf{(Müller et al., 2006)} present how procedural model generation is accomplished using Shape Grammar: a method developed by \textbf{(Stiny, 1975)} for automatically and iteratively constructing 3D models based in a hierarchical set of rules.
Shape grammar was originally developed from a method known as L-systems: a system for hierarchical model description which was originally used to generate plants for 3D animation \textbf{(Parish and Müller, 2001)}.
Shape grammar works by starting with a crude volumetric model of a building and iteratively remodeling it from a set of shape grammars.
Shape grammars often work from a data bank of 3D models and a set of hierarchical rules for how these can be pieced together.
This creates a possible generative space of many different models that can be created from the same data and rule set, and with the induction of random fluctuations, many different models can be generated from the same dataset and rules.
Using this procedure, a procedural model can structure the façade of a building, add texture and add minor detail models such as windows, doors and ornaments, and so on.
The maximum accuracy that can be obtained in these finer details is largely dependent on the availability of such data, and as most commercially available GIS data only serve crude building footprints and road placement, the model will often have to make guesses about the finer detail, which sacrifices accuracy to the real world.

@@

Space syntax, developed by \textbf{(Hillier, 1997)}, is a set of theories that analyze complex, large-scale spaces such as cities, and tries to explain human behavior from a spatial point of view.
It is commonly used in the analysis of procedural generation models of urban environments.
The central idea is that spatial structures such as building patterns both reflect and create the patterns for human interaction with their surroundings.
A common theme in space syntax is the division of spacial structures into different levels of observation, and then identifying different units of space and the pattern in which they connect to each other.
Space syntax is an analytical method that relies on the availability of city maps, which has since its invention been developed into an extensive research fields on its own
\textbf{(Parish and Müller, 2001)}.
It has also been widely used in the analysis of pedestrian flows and way finding
\textbf{(Peponis, Zimring and Choi, 1990)}.
Furthermore, the idea of using space syntax to divide spatial structures into different levels of observation is frequently used in the research field of urban design and urban planning, to gain an understanding how space is utilized and connected at different levels of usage.
One can for example analyze a city on an interior-by-interior basis, or use city blocks and streets as the least unit of space, depending on the goal of the study.
\textbf{(Stojanovski et al., 2020)}.

@@

To assess the completeness of their OSM data, the study by \textbf{(Fan et al., 2014)} used the fraction of the total building area in the reference dataset and the OSM dataset.
The motivation for this was to eliminate semantic error, which is a type of systematic error due to differences in how data is presented in both datasets.
The most common such difference is the fact that a large building in one dataset may be segmented into several smaller ones in the other.
Because of this, a building in one set might be represented as an aggregation of multiple buildings in the other set (see figure \ref{fig:building-match-types} for a visual example).
This makes it impossible to find a complete one-to-one building footprint mapping between the two datasets, and thus the idea is that using building area as an estimator for completeness will be less sensitive to semantic errors than using the number of buildings or other objects in each dataset.

\textbf{(Fan et al., 2014)} further discussed how the relative overlap in building area can be used as an estimate of the \textbf{building correspondence} between datasets.
Building correspondence is a term that refers to how well it can be judged which building footprints between two datasets that correspond to the same building in real life.
Using the relative area overlap between buildings to find which footprints represent the same building can only be done in cases where there is not much displacement between OSM building footprints data and the reference data set.
The relative building overlap between the OSM and the reference datasets is defined as follows, given the footprints of any building in the OSM set ($foot_{OSM}$) and the reference set ($foot_{REF}$):
\begin{center}
    $S_{RO}(foot_{OSM}, foot_{REF}) = \frac{A_{Overlap}}{min(A_{foot_{OSM}}, A_{foot_{REF}}))}$
\end{center}

@@

Once the building correspondence has been found and a building footprint mapping has been established (where any building in one dataset points to one, more or no buildings in the other dataset) the next challenge is evaluating how similar two matching building footprints are.
When assessing the positional accuracy between the points of two building footprint polygons, first one must find a matching set of vertices between a pair of matching buildings, which is not always trivial since the vertex counts of the two buildings may not be the same.
The two buildings might have been modeled at different levels of detail, or vertex clusters may be found at different parts of the polygon in the two datasets.
To address this problem, the OSM data precision study by \textbf{(Fan et al., 2014)} used the Douglas-Peucker algorithm, first defined by \textbf{(Douglas and Peucker, 1973)}, to reduce both building polygons to a similar vertex count, and then proceed with locating the most likely matching vertices in the polygons.
The full procedure for matching points will be presented in section 3.2.3 of this report.

The study by \textbf{(Fan et al., 2014)} also obtained a measure of the \textbf{shape accuracy} of building footprints.
While the positional accuracy accounts for the error between individual points in a polygon, the shape accuracy is a measure of how similar two polygons at a higher level of observation, regardless of relative scaling and rotation of the two polygons.
The shape accuracy between two matching buildings was quantified by using a similarity function that depends on the difference between the turning functions of matching building footprints in both datasets.
The turning function was first defined by \textbf{(Arkin et al., 1991)}, as a method for measuring the similarity of two polygons, and measures the cumulative angle of a single polygon's counter-clockwise tangent, as a function of the cumulative normalized length when traveling along the edge of the polygon.
The turning function effectively steps over the tangent of a polygon continuously for a full cycle, and logs the total turning angle by adding together the angle of each turn.
It includes consideration of whether the individual turns are clockwise or counter-clockwise, and is invariant to rotation and scaling of the polygon.
Once the turning functions of two similar polygons have been obtained, the integral between them is taken as a measure of shape accuracy.
The full procedure for calculating shape accuracy will be presented in section 3.2.2 of this report.

@@

Another crucial concept in city modeling and GIS mapping is the Level Of Detail (LOD) aspect.
The LOD defines a number of possible different representation of the same GIS data, which vary in the amount of detail in its presentation.
A modeler can choose to present the same data (e.g. building, roads, or decorative models) at different detail levels, depending on the use and application.
Furthermore, the LOD chosen for a particularly application also suggests how the data has been acquired and modeled.
It turns out however, in practice, that the terminology and use of LOD is still ambiguous and the definitions vary greatly between practitioners, standards and institutions, despite its usefulness.
\textbf{(Biljecki et al., 2014)}

@@

The methods used for the geodata precision study were largely based on those used by \textbf{(Fan et al., 2014)} in their study of the quality of the OSM Munich dataslice in 2014.
First a feature map was found, which mapped individual buildings between the OSM dataset and a reference dataset.
A program then took matching building pairs and extracted matching geometrical points after first simplifying the polygons using the Douglas-Peucker algorithm by \textbf{(Douglas and Peucker, 1973)}.
Once a set of matching points has been found across the whole dataset, the average distance between them was taken as a measure of the positional accuracy.

Additionally, the completeness of the OSM dataset (a measure of how many features are represented in both the OSM dataset and the reference dataset) was calculated as the relative difference in area of all building footprints, between the two datasets.
Finally, the shape accuracy of the OSM dataset has also been calculated and graphed as an extra verification of the similarity between the datasets.
